Notes on filtration of logics containing {\bf K5}

### 2021-02-27 filtration-stuff/Text.tex
refactor proof

If $\Phi$ is finite subset of $\Gamma$ and $\sim = \sim_{\Phi}$, then $\widehat{\mathcal{M}}$ is a definable $\Gamma$-filtration of $\mathcal{M}$ through $\Phi$. If $\sim = \sim_{\Gamma}$, then such a filtration by means of the definiton above is called \emph{strict}.

@@

A subset $\Delta \subseteq \Gamma$ is a \emph{finite $\mathcal{L}$-theory} if $\bigwedge \Delta \notin \mathcal{L}$. A subset $\Delta$ is maximal, if (the following are obviouslyequivalent):
\begin{enumerate}
  \item $\Delta$ is maximal amongst finite $\mathcal{L}$-theories,
  \item For each $\psi \in \operatorname{Sub}(\varphi)$ either $\psi \in \Delta$ or
  $\neg \psi \in \Delta$.
\end{enumerate}

@@

\begin{defin}
  Let $\mathcal{L}$ be a normal modal logic and $\varphi \not\in \mathcal{L}$. A finite ``canonical'' model is a triple $\mathcal{M}_{\mathcal{L}}^{\varphi} = \langle W_{\mathcal{L}}^{\varphi}, R_{\mathcal{L}}^{\varphi}, \vartheta_{\mathcal{L}}^{\varphi} \rangle$, where
  \begin{enumerate}
    \item $W_{\mathcal{L}}^{\varphi}$ is the set all maximal theories that extend finite $\mathcal{L}$-theories
    \item $R_{\mathcal{L}}^{\varphi}$ is a relation such that
    $\langle W_{\mathcal{L}}^{\varphi}, R_{\mathcal{L}}^{\varphi} \rangle$ is an $\mathcal{L}$-frame
    and
    \begin{center}
    $\forall \Box \psi \in \operatorname{Sub}(\varphi) \:\:
    \forall \Delta_1 \in W_{\mathcal{L}}^{\varphi} \:\: (\Box \psi \in \Delta_1 \Leftrightarrow
    \forall \Delta_2 \in R_{\mathcal{L}}^{\varphi}(\Delta_1) \:\: \psi \in \Delta_2)$
    \end{center}
    \item $\vartheta_{\mathcal{L}}^{\varphi}(p) = \{ \Delta \in W_{\mathcal{L}}^{\varphi} \: | \: p \in \Delta \}$.
  \end{enumerate}
\end{defin}

\begin{theorem} Let $\mathcal{L}$ be a canonical logic, then the following are equivalent:
  \begin{enumerate}
    \item $\mathcal{L}$ admits strict filtrations
    \item If $\varphi \not\in \mathcal{L}$, there exists a finite ``canonical'' model $\mathcal{M}_{\mathcal{L}}^{\varphi}$ such that
    $\mathcal{M}_{\mathcal{L}}^{\varphi} \models \mathcal{L}$ and
    $\mathcal{M}_{\mathcal{L}}^{\varphi}$ refutes $\varphi$.
  \end{enumerate}
\end{theorem}

@@

  \begin{enumerate}
    \item $(\Rightarrow)$

    Let $\varphi \not\in \mathcal{L}$, there exists a model $\mathcal{M}$ that refutes $\varphi$, so there exists $w \in \underline{\mathcal{M}}$ such that $\mathcal{M}, w \nvDash \varphi$. But $\mathcal{L}$ admits strict filtration, so there exists a model
    $\widehat{\mathcal{M}} = \langle \widehat{W}, \widehat{R}, \widehat{\vartheta} \rangle$, where $\widehat{W} = W / \sim_{\varphi}$ and $\hat{R}^{min} \subseteq \widehat{R} \subseteq \hat{R}^{max}$. In particular, $\widehat{\mathcal{M}}, \widehat{w} \nvDash \varphi$.
    Let us put $T = \operatorname{Th}(\widehat{\mathcal{M}})$.

    ...

    On the other hand, $\mathcal{L}$ is canonical. Therefore, $\mathcal{F}_{\mathcal{L}} \models \mathcal{L}$ as well as $\mathcal{F}_{T} \models \mathcal{L}$.

    Here $\mathcal{F}_{T}$ is a frame $\langle W_T, R_T \rangle$, where $W_T$ is the set of all maximal consistent theories that contain $T$ and $R_T = R_{\mathcal{L}} \cap W_T \times W_T$.

    $\mathcal{M}_{T} = \langle \mathcal{F}_{\widehat{\mathcal{M}}}, \vartheta_{\widehat{\mathcal{M}}} \rangle$,
    being a generated submodel of $\mathcal{M}_{\mathcal{L}}$, does not have to be a ``finite canonical'' model according to our definition.

    Let us put $W_{\mathcal{L}}^{\varphi} = \{ \Gamma \subseteq \operatorname{Sub}(\varphi) \cup \neg \operatorname{Sub}(\varphi) \: | \: \Gamma \text{ is maximal consistent} \}$.

    ...

@@

    Let $[u] \in W/\sim_{\operatorname{Sub}(\varphi)}$, where $W = \underline{\mathcal{M}}$,
    consider its truth set $||[u]|| = \{ \psi \in \operatorname{Sub}(\varphi) \: | \: \mathcal{M}, u \models \psi \}$.
    $||[u]||$ is clearly a finite consistent theory and it consists in some $\Delta \in W_{\mathcal{L}}^{\varphi}$ by the finite analogue of Lindenbaum's lemma. Such a $\Delta$ is clearly unique as well.
    Consider a map
    \begin{center}
    $\iota : W/\sim_{\operatorname{Sub}(\varphi)} \to W_{\mathcal{L}}^{\varphi}$ with
    $\iota : [u] \mapsto \Delta \supseteq ||[u]||$
    \end{center}

    Consider the set $\iota(W/\sim_{\operatorname{Sub}(\varphi)})$ denoted as $\operatorname{Im}(\iota)$. Let us show that $\operatorname{Im}(\iota)$ is $R_{\mathcal{L}}^{\varphi}$-closed.

@@

    Consider $\Delta$ as a finite ultrafilter. Thus, $\Delta$ is principal and it 's generated by $\psi_1 \land \dots \land \psi_n$, where each $\psi_i \in \operatorname{Sub}(\varphi)$. So $\Diamond (\psi_1 \land \dots \land \psi_n) \in \iota([u])$.
    Then $\mathcal{M}, u \models \Diamond (\psi_1 \land \dots \land \psi_n)$. That implies $\mathcal{M}, v \models \psi_1 \land \dots \land \psi_n$ for some $v \in R(u)$.
    This is clearly a desired $v$ since $||[v]|| \subseteq \Delta$ and such a $\Delta$ is unique.

    Thus, $\langle \operatorname{Im}(\iota), R_{\mathcal{L}}^{\varphi} \cap \operatorname{Im}(\iota) \times \operatorname{Im}(\iota), {\vartheta'}_{\mathcal{L}}^{\varphi} \rangle$, where ${\vartheta'}_{\mathcal{L}}^{\varphi}(p) = \vartheta_{\mathcal{L}}^{\varphi}(p) \cap \operatorname{Im}(\iota)$ is a generated submodel of $\mathcal{M}_{\mathcal{L}}^{\varphi}$.

    Consider a model $\widehat{\mathcal{M}} = \langle W/\sim_{\operatorname{Sub}(\varphi)}, \widehat{R}, \widehat{\vartheta} \rangle$, where $\widehat{R}$ is defined as
    \begin{center}
      $[u] \widehat{R} [v]$ iff $||[u]|| {R'}_{\mathcal{L}}^{\varphi} ||[v]||$, where ${R'}_{\mathcal{L}}^{\varphi} R_{\mathcal{L}}^{\varphi} \cap \operatorname{Im}(\iota) \times \operatorname{Im}(\iota)$.
    \end{center}
    Let us prove that $\widehat{\mathcal{M}}$ is a filtration of $\mathcal{M}$ through $\operatorname{Sub}(\varphi)$.

@@

    Suppose that there are $u' \in [u]$ and $v' \in [v]$ with $u' R v'$. Let us show that
    $[u] \widehat{R} [v]$. That is, let us show that $\Diamond \iota(v) \subseteq \iota(v)$.
    But if $\mathcal{M}, v' \models \psi$ for $\Diamond \psi \in \operatorname{Sub}(\varphi)$, then
    $\mathcal{M}, u' \models \Diamond \psi$. Recall that $u' \sim_{\operatorname{Sub}(\varphi)} u$ and $v' \sim_{\operatorname{Sub}(\varphi)} v$, so we have $\Diamond ||[v]|| \subseteq ||[u]||$.
    The same inclusion clearly holds for the maximal extensions of $\Diamond ||[v]||$ and $||[u]||$.

>>>>

If $\Phi$ is finite subset of $\Gamma$ and $\sim = \sim_{\Phi}$, then $\widehat{\mathcal{M}}$ is a definable $\Gamma$-filtration of $\mathcal{M}$ through $\Phi$. If $\sim = \sim_{\Gamma}$, then such a filtration by means of the definiton above is called \emph{strict}. A class of models $\mathbb{M}$ admits strict filtrations for models (ASF), if for every $\operatorname{Sub}$-closed set $\Gamma$ and for every $\mathcal{M} \in \mathbb{M}$ there exists a $\Gamma$ filtration of $\mathcal{M}$.
A class of frames $\mathbb{F}$ admits strict filtrations for frames, if for every $\operatorname{Sub}$-closed set $\Gamma$ and for every frame $\mathcal{F} \in \mathbb{F}$ and every model $\mathcal{M}$ over $\mathcal{F}$ there exists a $\Gamma$ filtration of $\mathcal{M}$. If $\mathcal{L}$ is canonical, then the ASF property for frames and ASF property for models are equivalent \cite[Theorem 2.10]{kikot2020completeness}.

@@

A subset $\Delta \subseteq \Gamma$ is a \emph{finite $\mathcal{L}$-consistent set} if $\neg \bigwedge \Delta \notin \mathcal{L}$. A subset $\Delta$ is maximal, if (the following are obviously equivalent):
\begin{enumerate}
  \item $\Delta$ is maximal amongst finite $\mathcal{L}$-consistent sets,
  \item For each $\psi \in \operatorname{Sub}(\varphi)$ either $\psi \in \Delta$ or
  $\neg \psi \in \Delta$.
\end{enumerate}

@@

\begin{defin}\label{fin}
  Let $\mathcal{L}$ be a normal modal logic and $\varphi \not\in \mathcal{L}$. A finite ``canonical'' model is a triple $\mathcal{M}_{\mathcal{L}}^{\varphi} = \langle W_{\mathcal{L}}^{\varphi}, R_{\mathcal{L}}^{\varphi}, \vartheta_{\mathcal{L}}^{\varphi} \rangle$, where
  \begin{enumerate}
    \item $W_{\mathcal{L}}^{\varphi}$ is the set all maximal theories that extend finite $\mathcal{L}$-theories
    \item $R_{\mathcal{L}}^{\varphi}$ is a relation such that
    $\langle W_{\mathcal{L}}^{\varphi}, R_{\mathcal{L}}^{\varphi} \rangle$ is an $\mathcal{L}$-frame
    and
    \begin{center}
    $\forall \Box \psi \in \operatorname{Sub}(\varphi) \:\:
    \forall \Delta_1 \in W_{\mathcal{L}}^{\varphi} \:\: (\Box \psi \in \Delta_1 \Leftrightarrow
    \forall \Delta_2 \in R_{\mathcal{L}}^{\varphi}(\Delta_1) \:\: \psi \in \Delta_2)$
    \end{center}
    \item $\vartheta_{\mathcal{L}}^{\varphi}(p) = \{ \Delta \in W_{\mathcal{L}}^{\varphi} \: | \: p \in \Delta \}$.
  \end{enumerate}
\end{defin}

\begin{theorem} $\mathcal{L}$ be a modal logic, then the following are equivalent:
  \begin{enumerate}
    \item $\mathcal{L}$ admits strict filtrations
    \item If $\varphi \not\in \mathcal{L}$, there exists a finite ``canonical'' model $\mathcal{M}_{\mathcal{L}}^{\varphi}$ such that
    $\mathcal{M}_{\mathcal{L}}^{\varphi} \models \mathcal{L}$ and
    $\mathcal{M}_{\mathcal{L}}^{\varphi}$ refutes $\varphi$.
  \end{enumerate}
\end{theorem}

@@

  \begin{enumerate}
    \item $(\Rightarrow)$
    Let $\varphi \not\in \mathcal{L}$, let us put $\Gamma = \operatorname{Sub}(\varphi)$.
    $\mathcal{L}$.
    In particular, $\mathcal{M}_{\mathcal{L}} \nvDash \varphi$, where $\mathcal{M}_{\mathcal{L}}$ is the canonical model of $\mathcal{L}$. $\mathcal{L}$ admits strict filtrations, so the filtration of $\mathcal{M}_{\mathcal{L}}$ through $\Gamma$ is also an $\mathcal{L}$-model.
    The underlying set of $\mathcal{M}_{\mathcal{L}} / \sim_{\Gamma}$ consists of maximal $\mathcal{L}$ theories up to $\Gamma$-equivalence and this quotient set is finite.

    It is readily checked that the quotient model $\mathcal{M}_{\mathcal{L}} / \sim_{\Gamma}$ satifsfies Definition~\ref{fin}.

@@

    Let $[u] \in W/\sim_{\operatorname{Sub}(\varphi)}$, where $W = \underline{\mathcal{M}}$,
    consider its truth set $||[u]|| = \{ \psi \in \operatorname{Sub}(\varphi) \: | \: \mathcal{M}, u \models \psi \}$.
    $||[u]||$ is clearly a finite $\mathcal{L}$-consistent set and it is maximal.
    Consider a map
    \begin{center}
    $\iota : W/\sim_{\operatorname{Sub}(\varphi)} \to W_{\mathcal{L}}^{\varphi}$ with
    $\iota : [u] \mapsto ||[u]||$
    \end{center}

    Consider the set $\iota(W/\sim_{\operatorname{Sub}(\varphi)})$ that we denote as $\operatorname{Im}(\iota)$. Let us show that $\operatorname{Im}(\iota)$ is $R_{\mathcal{L}}^{\varphi}$-closed.

@@

    Note that $\bigwedge \limits_{\psi \in \Delta} \psi \in \Delta$ since $\Delta$ is finite.
    This conjuction has the form $\bigwedge \limits_{\psi \in \Delta} \psi \in \Delta = \psi_1 \land \dots \land \psi_n$ for some $n < \omega$ and for some $\psi_i \in \operatorname{Sub}(\varphi) \cup \neg \operatorname{Sub}(\varphi)$, where $0 < i < \omega$.

    Therefore $\mathcal{M}^{\varphi}_{\mathcal{L}}, \iota([u]) \models \Diamond (\psi_1 \land \dots \land \psi_n)$.
    Then $\mathcal{M}, u \models \Diamond (\psi_1 \land \dots \land \psi_n)$. That implies $\mathcal{M}, v \models \psi_1 \land \dots \land \psi_n$ for some $v \in R(u)$.
    This is clearly a desired $v$ since $||[v]|| \subseteq \Delta$ and such a $\Delta$ is unique.

    Thus, $\langle \operatorname{Im}(\iota), R_{\mathcal{L}}^{\varphi}, {\vartheta'}_{\mathcal{L}}^{\varphi} \rangle$, where ${\vartheta'}_{\mathcal{L}}^{\varphi}(p) = \vartheta_{\mathcal{L}}^{\varphi}(p) \cap \operatorname{Im}(\iota)$
    is a generated submodel of $\mathcal{M}_{\mathcal{L}}^{\varphi}$.

    Consider a model $\widehat{\mathcal{M}} = \langle W/\sim_{\operatorname{Sub}(\varphi)}, \widehat{R}, \widehat{\vartheta} \rangle$, where $\widehat{R}$ is defined as
    \begin{center}
      $[u] \widehat{R} [v]$ iff $||[u]|| {R}_{\mathcal{L}}^{\varphi} ||[v]||$
    \end{center}

@@

    Suppose that there are $u' \in [u]$ and $v' \in [v]$ with $u' R v'$. Let us show that
    $[u] \widehat{R} [v]$. That is, let us show that $\Diamond \iota(v) \subseteq \iota(v)$.
    But if $\mathcal{M}, v' \models \psi$ for $\Diamond \psi \in \operatorname{Sub}(\varphi)$, then
    $\mathcal{M}, u' \models \Diamond \psi$. Recall that $u' \sim_{\operatorname{Sub}(\varphi)} u$ and $v' \sim_{\operatorname{Sub}(\varphi)} v$, so we have $\Diamond ||[v]|| \subseteq ||[u]||$.
